---
title: "Codebook"
author: "Jacob Álvarez"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
IC <- read_csv('bia_indian_lands_dataset.csv')

#rm unnecessary variables
IC <-
  IC |>
    select(-fid, -objectid, -area, -perimeter, -aiana,
           -ind_source, -ind_scale, -org_code, -region_cod,
           -agency_nam, -osgt_code, -osgt_assoc, -native_cor,
           -mail_name, -cnty_st, -inc_percap, -tribal_num,
           -treaty, -tribal_est, -gov_type, -acres, -agency_cod,
           -poi_on_ima, -field_veri, -shape_leng, -shape_le_1,
           -shape_le_2, -irmp_plan, -short_name, -shape_le_3,
           -shape_area, -air_020107, -air_20108, -r_code,
           -ind_name, -fed_regist, -labor_tot)


#rm observations when not associated with a native entity
IC <-
  IC[!is.na(IC$entity),]

##factorize and reorder aiana_desc

IC$aiana_desc = 
  as_factor(IC$aiana_desc)

fct_count(IC$aiana_desc)

IC <-
  IC |>
  mutate(
    aiana_desc = fct_relevel(
      aiana_desc, # source variable
      'Trust land for which no reservation exists',
      'Federally Recognized Tribal Entity',
      'Trust Land related to a Federally Recognized Reservation',
      'Alaska Native Village',
      'American Indian Reservation'
    )
  )    

levels(IC$aiana_desc)

IC$self_gover = 
  as_factor(IC$self_gover)

fct_count(IC$self_gover)

IC$gaming = 
  as_factor(IC$gaming)

fct_count(IC$gaming)

IC$med_fac = 
  as_factor(IC$med_fac)

fct_count(IC$med_fac)
##med_fac to binary

IC <-
IC|>
mutate(
    med_fac = case_when(grepl("none", med_fac) ~ "NO",
                        grepl("clinic", med_fac, ignore.case = TRUE) ~"YES"))
    

##area pcts

IC <-
  IC |>
    mutate(tribe_area_pct = tribe_area / tot_area * 100,
           fed_area_pct = fed_area / tot_area * 100,
           allot_area_pct = allot_area / tot_area * 100)

IC <-
  IC |>
  select(-tribe_area, - allot_area, -fed_area, -tot_area)

##rm NaN, Inf

IC <-
  IC |>
  mutate(
    tribe_area_pct = as.numeric(tribe_area_pct),
    tribe_area_pct = case_when(
      tribe_area_pct == 'NaN' ~ NA, 
      tribe_area_pct == 'Inf' ~ 100,
      TRUE ~ as.numeric(tribe_area_pct)
      )
  )

IC <-
  IC |>
  mutate(
    fed_area_pct = as.numeric(fed_area_pct),
    fed_area_pct = case_when(
      fed_area_pct == 'NaN' ~ NA,
      tribe_area_pct == 'Inf' ~ 100,
      TRUE ~ as.numeric(fed_area_pct)
    )
  )

IC <-
  IC |>
  mutate(
    allot_area_pct = as.numeric(allot_area_pct),
    allot_area_pct = case_when(
      allot_area_pct == 'NaN' ~ NA,
      tribe_area_pct == 'Inf' ~ 100,
      TRUE ~ as.numeric(allot_area_pct)
    )
  )

##pop per sq mi

IC <-
  IC |>
    mutate(sq_miles = round(sq_miles, 1))

IC <-
  IC |>
    mutate(tribe_pop_sq_mi = pop_tot / sq_miles)

IC <-
  IC |>
    mutate(tribe_pop_sq_mi = round(tribe_pop_sq_mi, 1),
           tribe_area_pct = round(tribe_area_pct, 1),
           fed_area_pct = round(fed_area_pct, 1),
           allot_area_pct = round(allot_area_pct, 1))

##mutate the economy

IC$econ_sourc <- stringr::word(IC$econ_sourc, 1)

#standardize obs names

IC$econ_sourc <- str_replace_all(IC$econ_sourc,
                                  c("," = "",
                                    ";" = "",
                                    "Fisheries" = "Fish",
                                    "Fishing" = "Fish",
                                    "Wild" = "Wild rice",
                                    "Retail/service" = "Retail"))
```

# Overview of Data

This dataset originally comes from the dataset available at [https://data.world/dhs/bia-indian-lands-dataset](https://data.world/dhs/bia-indian-lands-dataset).

# Sources and Methodology

This data was compiled by the Department of Homeland Security's Homeland Infrastructure Foundation and made available to the public in 2016. The modified dataset that I have created consists of 696 observations of 19 variables that give a snapshot of tribal lands in the United States in the late 2010s.

# Missing Values

All missing values are designated with `NA`.

# Itemized Presentation of Variables

## Land type
Variable Name: `aiana_desc`

Variable Type: factor

Description: Legal classification of land
```{r, echo = F}
eval = T
knitr::kable(fct_count(IC$aiana_desc))
```

## Entity

Variable Name: `entity`

Variable Type: character

Description: Name of tribal entity that land is associated with

## Region Name

Variable Name: `region_nam`

Variable Type: character

Description: US region that land is in

## Self Government

Variable Name: `self_gover`

Variable Type: factor

Description: Whether the land is under tribal self government
```{r, echo = F}
eval = T
knitr::kable(fct_count(IC$self_gover))
```

## State

Variable Name: `state`

Variable Type: character

Description: US state that land is in, 2-letter abbreviation

## High School Graduation

Variable Name: `hsgrad_pct`

Variable Type: numeric

Description: Percent of tribal members who have graduated high school
```{r, echo = F}
eval = T
summary(IC$hsgrad_pct)
```

## College Graduation

Variable Name: `colgrad_pct`

Variable Type: numeric

Description: Percent of tribal members who have graduated college
```{r, echo = F}
eval = T
summary(IC$colgrad_pc)
```

## Unemployment Percent

Variable Name: `unemploy_p`

Variable Type: numeric

Description: Percent of tribal members who are unemployed
```{r, echo = F}
eval = T
summary(IC$unemploy_p)
```

## Total Population

Variable Name: `pop_tot`

Variable Type: numeric

Description: Number of tribal members
```{r, echo = F}
eval = T
summary(IC$pop_tot)
```

## Main Economic Activity

Variable Name: `econ_sourc`

Variable Type: character

Description: Main economic activity that tribe is engaged in

## Operates Casinos

Variable Name: `gaming`

Variable Type: factor

Description: Whether the tribe has gambling operations
```{r, echo = F}
eval = T
knitr::kable(fct_count(IC$gaming))
```

## Medical Facilities

Variable Name: `med_fac`

Variable Type: factor

Description: Whether the land has a medical facility

```{r, echo = F}
eval = T
knitr::kable(fct_count(IC$med_fac))
```

## Area

Variable Name: `sq_miles`

Variable Type: numeric

Description: Area of land in square miles
```{r, echo = F}
eval = T
summary(IC$sq_miles)
```

## Tribe Area Percent

Variable Name: `tribe_area_pct`

Variable Type: numeric

Description: What percent of the land is tribal land
```{r, echo = F}
eval = T
summary(IC$tribe_area_pct)
```

## Federal Area Percent

Variable Name: `fed_area_pct`

Variable Type: numeric

Description: What percent of the land is federal land
```{r, echo = F}
eval = T
summary(IC$fed_area_pct)
```

## Allotment Area Percent

Variable Name: `allot_area_pct`

Variable Type: numeric

Description: What percent of the land is allotment land (private property)
```{r, echo = F}
eval = T
summary(IC$allot_area_pct)
```

## Population per Square Mile

Variable Name: `tribe_pop_sq_mi`

Variable Type: numeric

Description: Tribal population per square mile of land
```{r, echo = F}
eval = T
summary(IC$tribe_pop_sq_mi)
```








